Lattices of acyclic pipe dreams

### 2025-02-18 acyclicFacets/draft/latticePipeDreams.tex
affiliation and grants

\author[V.~Pilaud]{Vincent Pilaud}
\address[V.~Pilaud]{Universitat de Barcelona}
\email{vincent.pilaud@ub.edu}
\urladdr{https://www.ub.edu/comb/vincentpilaud/}

\thanks{
NB was supported by NSERC and York Research Chair in Applied Algebra.
CC was supported by the Austrian Science Fund FWF (Project P 33278).
NC \& VP were partially supported by the French project CHARMS (ANR~19\,CE40\,0017).
CC, NC \& VP were also supported by the Austrian\,--\,French project PAGCAP (ANR~21\,CE48\,0020 \& FWF I 5788).
VP was a CNRS researcher at \'Ecole Polytechnique when this work was done.
\\
\hspace*{.5cm} 
Part of the material of this paper was announced in an extended abstract of the conference FPSAC'23~\cite{Cartier-FPSAC}.
}

@@

\vspace*{-1.2cm}

@@

\vspace*{-.7cm}

@@

\section{Introduction}
\label{sec:introduction}

@@

\begin{figure}[t]
	\centerline{
		%\input{pipeDreams}
		\includegraphics[scale=.9]{pipeDreams}
	}
	\caption{Two pipe dreams of~$\pipeDreams(1365724)$ connected by an increasing flip (exchanging a contact with the crossing on the two red pipes~$4$ and~$5$).}
	\label{fig:pipeDreams}
\end{figure}

@@

Note that the pipe dream congruence indeed defines an equivalence relation by \cref{thm:partitionPipeDreams}.
In this section, we prove that it is a lattice congruence.

@@

\pagebreak
\begin{lemma}
\label{lem:linearExtensionsGreedyFacets}
Let $\greedyFacet$ and $\antiGreedyFacet$ be the greedy and antigreedy facets of $\subwordComplex(Q,\omega)$, respectively.
Then 
\begin{enumerate}
\item $e\in \linearExtensions(\greedyFacet)$.
\item $\omega \in \linearExtensions(\antiGreedyFacet)$.
\item If $e\in \linearExtensions(I)$ then $I=\greedyFacet$.
\end{enumerate}
\end{lemma}

@@

It starts by setting $I^0=\varnothing$ to be the empty set.
It then scans the word $Q=(q_1,\dots ,q_m)$ from left to right.
At position $j\in [m]$ it produces a new set $I^j$ obtained from the previous one $I^{j-1}$ by either adding $j$ or not, according to certain rules.
Roughly speaking, the goal of the algorithm is to insert a reduced expression of $\omega$ in $Q$ by sweeping the word from left to right, while deciding whether we use a letter or not by looking at the root $\beta=\rootFunction{I^j}{j}$ that it produces.
If $\beta \in \Ninv(\omega)$ then it can not be used (Case~\ref{sweeping_case1}), otherwise we would not have a reduced expression for $\omega$.
If $\beta \in \Inv(\omega)$ then it takes it when $\beta \in \Inv(\pi)$ (Case~\ref{sweeping_case2}), or delays it as much as possible when $\beta \in \Ninv(\pi)$ (Case~\ref{sweeping_case3}).
If $\beta \in -\Inv(\omega)$ it can simply not take it (Case~\ref{sweeping_case4}).

>>>>

\author[V.~Pilaud]{Vincent Pilaud}
\address[Vincent Pilaud]{Universitat de Barcelona \& Centre de Recerca Matemàtica, Barcelona, Spain}
\email{vincent.pilaud@ub.edu}
\urladdr{https://www.ub.edu/comb/vincentpilaud/}

\thanks{
NB was supported by NSERC and York Research Chair in Applied Algebra.
CC was supported by the Austrian Science Fund FWF (Project P 33278).
CC, NC \& VP were supported by the Austrian\,--\,French project PAGCAP (ANR~21\,CE48\,0020 \& FWF I 5788).
NC \& VP were also supported by the French project CHARMS (ANR~19\,CE40\,0017).
VP was also supported by the Spanish project PID2022-137283NB-C21 of MCIN/AEI/10.13039/501100011033 / FEDER, UE, by the Spanish--German project COMPOTE (AEI PCI2024-155081-2 \& DFG 541393733), by the Severo Ochoa and María de Maeztu Program for Centers and Units of Excellence in R\&D (CEX2020-001084-M), by the Departament de Recerca i Universitats de la Generalitat de Catalunya (2021 SGR 00697).
VP was a CNRS researcher at \'Ecole Polytechnique when this work was done.
\\
\hspace*{.5cm} 
Part of the material of this paper was announced in an extended abstract of the conference FPSAC'23~\cite{Cartier-FPSAC}.
}

@@

%\vspace*{-1.2cm}

@@

%\vspace*{-.7cm}

@@

\pagebreak
\section{Introduction}
\label{sec:introduction}

@@

\begin{figure}[b]
	\centerline{
		%\input{pipeDreams}
		\includegraphics[scale=.9]{pipeDreams}
	}
	\caption{Two pipe dreams of~$\pipeDreams(1365724)$ connected by an increasing flip (exchanging a contact with the crossing on the two red pipes~$4$ and~$5$).}
	\label{fig:pipeDreams}
\end{figure}

@@

\pagebreak
Note that the pipe dream congruence indeed defines an equivalence relation by \cref{thm:partitionPipeDreams}.
In this section, we prove that it is a lattice congruence.

@@

\begin{lemma}
\label{lem:linearExtensionsGreedyFacets}
Let $\greedyFacet$ and $\antiGreedyFacet$ be the greedy and antigreedy facets of $\subwordComplex(Q,\omega)$, respectively.
Then 
\begin{enumerate}
\item $e\in \linearExtensions(\greedyFacet)$.
\item $\omega \in \linearExtensions(\antiGreedyFacet)$.
\item If $e\in \linearExtensions(I)$ then $I=\greedyFacet$.
\end{enumerate}
\end{lemma}

@@

It starts by setting $I^0=\varnothing$ to be the empty set.
It then scans the word $Q=(q_1,\dots ,q_m)$ from left to right.
At position $j\in [m]$ it produces a new set $I^j$ obtained from the previous one~$I^{j-1}$ by either adding $j$ or not, according to certain rules.
Roughly speaking, the goal of the algorithm is to insert a reduced expression of $\omega$ in $Q$ by sweeping the word from left to right, while deciding whether we use a letter or not by looking at the root $\beta=\rootFunction{I^j}{j}$ that it produces.
If~$\beta \in \Ninv(\omega)$ then it can not be used (Case~\ref{sweeping_case1}), otherwise we would not have a reduced expression for $\omega$.
If~$\beta \in \Inv(\omega)$ then it takes it when $\beta \in \Inv(\pi)$ (Case~\ref{sweeping_case2}), or delays it as much as possible when $\beta \in \Ninv(\pi)$ (Case~\ref{sweeping_case3}).
If $\beta \in -\Inv(\omega)$ it can simply not take it (Case~\ref{sweeping_case4}).
